---
title: "bank_dataset"
author: "Rosemary Saoke"
date: "11/02/2024"
output:
  html_document: default
  pdf_document: default
---



```{r setup, include=FALSE}
# Set the CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Install and load required packages
if (!require("readxl")) install.packages("readxl")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("dplyr")) install.packages("dplyr")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("psych")) install.packages("psych")

library(readxl)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(psych)
```


```{r bank dataset}
# Load the dataset
bank_dataset <- read_excel("~/Documents/demo datasets/bank_dataset.xlsx")
```
Brief description of the dataset

This dataset contains information about 10,001 customers of a bank. The data includes various demographic and financial attributes of the customers, such as their:
- Credit Score
- Customer ID
- Surname
- Geography (Country)
- Gender
- Age
- Tenure (Number of years with the bank)
- Balance (Account balance)
- Number of Products (Number of bank products the customer uses)
- Has Credit Card (Binary indicator if the customer has a credit card)
- Is Active Member (Binary indicator if the customer is an active member)
- Estimated Salary
- Exited (Binary indicator if the customer has exited the bank). 
This summary is based on the first 10,001 customers selected from the original dataset.

```{r credit score analysis}
# Histogram of Credit Score
ggplot(bank_dataset, aes(x = CreditScore)) +  
  geom_histogram(fill = "darkblue", color = "black", bins = 30) +  
  labs(title = "Distribution of Credit Score",
       x = "Credit Score",
       y = "Frequency") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Summary statistics for CreditScore variable
summary_stats_creditscore <- describe(bank_dataset$CreditScore)
summary_stats_creditscore
```

Conclusion: The distribution of credit scores appears to be somewhat normally distributed with a slight negative skew. The majority of customers have credit scores around the median value.

```{r age analysis}
# Histogram of Ages 
ggplot(bank_dataset, aes(x = Age)) +  
  geom_histogram(fill = "darkblue", color = "black", bins = 60) + 
  labs(title = "Distribution of Age",
       x = "Age",
       y = "Frequency") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Summary statistics for Age variable
summary_stats_age <- describe(bank_dataset$Age)
summary_stats_age
```

Conclusion: The age distribution is positively skewed, indicating that there are more younger customers in the dataset. The distribution is relatively uniform with a peak around the mid-30s.

```{r balance analysis}
# Histogram of Balances 
ggplot(bank_dataset, aes(x = Balance)) +  
  geom_histogram(fill = "darkblue", color = "black", bins = 50) +  
  labs(title = "Distribution of Balance",
       x = "Balance",
       y = "Frequency") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Summary statistics for Balance variable
summary_stats_balance <- describe(bank_dataset$Balance)
summary_stats_balance
```
Conclusion: Most customers seem to have zero balance, indicating that they might have multiple accounts or the dataset includes inactive accounts. For non-zero balances, there is a normal distribution with a mean around 55457.39.

```{r estimated salary analysis}
# Histogram of Estimated Salary 
ggplot(bank_dataset, aes(x = EstimatedSalary)) +  
  geom_histogram(fill = "darkblue", color = "black", bins = 75) +  
  labs(title = "Distribution of Estimated Salary",
       x = "Estimated Salary",
       y = "Frequency") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Summary statistics for Estimated Salary variable
summary_stats_estimatedsalary <- describe(bank_dataset$EstimatedSalary)
summary_stats_estimatedsalary
```

Conclusion: The distribution of estimated salaries appears to be slightly negatively skewed with the majority of salaries around the median value of 118324.41.


```{r heatmap}
# Heatmap of numerical values
numeric_vars <- bank_dataset[, c("CreditScore", "Age", "Tenure", "Balance", "NumOfProducts", "EstimatedSalary")]
correlation_matrix <- cor(numeric_vars)
correlation_df <- as.data.frame(correlation_matrix)
rownames(correlation_df) <- colnames(correlation_df) <- colnames(numeric_vars)
melted_correlation <- correlation_df %>%
  rownames_to_column(var = "variable") %>%
  gather(key = "variable2", value = "value", -variable)
ggplot(melted_correlation, aes(variable, variable2, fill = value, label = round(value, 2))) +
  geom_tile() +
  geom_text(color = "black", size = 3) +  
  scale_fill_gradient2(low = "green", mid = "white", high = "darkgreen", midpoint = 0, name = "Correlation") +
  labs(title = "Correlation Heatmap of Numerical Variables",
       x = "Variables",
       y = "Variables")
```

Conclusion:There is a moderate negative correlation between the number of products and balance, suggesting that customers with more products tend to have lower balances.
There are very weak negative correlations between the number of products and age, and very weak positive correlations between balance and age.

```{r boxplot:credit score and gender}
# Box plots
# Boxplot showing the distribution of credit score between genders
ggplot(bank_dataset, aes(x = Gender, y = CreditScore, fill = Gender)) +
  geom_boxplot() +
  scale_fill_manual(values = c("red", "darkblue")) +  
  labs(title = "Credit Score Distribution by Gender",
       x = "Gender",
       y = "Credit Score")

# Summary statistics for Credit Score by Gender
summary_male <- bank_dataset %>%
  filter(Gender == "Male") %>%
  summarise(
    Median = median(CreditScore, na.rm = TRUE),
    Q1 = quantile(CreditScore, 0.25, na.rm = TRUE),
    Q3 = quantile(CreditScore, 0.75, na.rm = TRUE),
    Min = min(CreditScore, na.rm = TRUE),
    Max = max(CreditScore, na.rm = TRUE)
  )
summary_female <- bank_dataset %>%
  filter(Gender == "Female") %>%
  summarise(
    Median = median(CreditScore, na.rm = TRUE),
    Q1 = quantile(CreditScore, 0.25, na.rm = TRUE),
    Q3 = quantile(CreditScore, 0.75, na.rm = TRUE),
    Min = min(CreditScore, na.rm = TRUE),
    Max = max(CreditScore, na.rm = TRUE)
  )

# Print summary statistics
cat("Summary Statistics for Males:\n")
print(summary_male)
cat("\nSummary Statistics for Females:\n")
print(summary_female)
```

Conclusion: The box plot shows the distribution of credit scores between genders. While there is some overlap, males tend to have slightly higher median credit scores compared to females.

```{r boxplot:credit score and geography}
# Boxplot showing the distribution of credit scores across different countries
ggplot(bank_dataset, aes(x = Geography, y = CreditScore, fill = Geography)) +
  geom_boxplot() +
  scale_fill_manual(values = c("darkblue", "darkgreen", "red")) + 
  labs(title = "Credit Score Distribution by Country",
       x = "Country",
       y = "Credit Score")

# Summary statistics for Credit Score by Geography 
summary_france <- bank_dataset %>%
  filter(Geography == "France") %>%
  summarise(
    Median = median(CreditScore, na.rm = TRUE),
    Q1 = quantile(CreditScore, 0.25, na.rm = TRUE),
    Q3 = quantile(CreditScore, 0.75, na.rm = TRUE),
    Min = min(CreditScore, na.rm = TRUE),
    Max = max(CreditScore, na.rm = TRUE)
  )

summary_germany <- bank_dataset %>%
  filter(Geography == "Germany") %>%
  summarise(
    Median = median(CreditScore, na.rm = TRUE),
    Q1 = quantile(CreditScore, 0.25, na.rm = TRUE),
    Q3 = quantile(CreditScore, 0.75, na.rm = TRUE),
    Min = min(CreditScore, na.rm = TRUE),
    Max = max(CreditScore, na.rm = TRUE)
  )

summary_spain <- bank_dataset %>%
  filter(Geography == "Spain") %>%
  summarise(
    Median = median(CreditScore, na.rm = TRUE),
    Q1 = quantile(CreditScore, 0.25, na.rm = TRUE),
    Q3 = quantile(CreditScore, 0.75, na.rm = TRUE),
    Min = min(CreditScore, na.rm = TRUE),
    Max = max(CreditScore, na.rm = TRUE)
  )

# Print summary statistics
cat("Summary Statistics for France:\n")
print(summary_france)
cat("\nSummary Statistics for Germany:\n")
print(summary_germany)
cat("\nSummary Statistics for Spain:\n")
print(summary_spain)
```

Conclusion: The box plot illustrates the distribution of credit scores from different countries. The median credit score seems to be slightly higher for French customers compared to German and Spanish customers.

```{r boxplot:estimated salary and gender}
# Boxplot showing the distribution of estimated salary between genders
ggplot(bank_dataset, aes(x = Gender, y = EstimatedSalary, fill = Gender)) +
  geom_boxplot() +
  scale_fill_manual(values = c("red", "darkblue")) +  
  labs(title = "Estimated Salary Distribution by Gender",
       x = "Gender",
       y = "Estimated Salary")

# Summary statistics for Estimated Salary by Gender
summary_male_salary <- bank_dataset %>%
  filter(Gender == "Male") %>%
  summarise(
    Median = median(EstimatedSalary, na.rm = TRUE),
    Q1 = quantile(EstimatedSalary, 0.25, na.rm = TRUE),
    Q3 = quantile(EstimatedSalary, 0.75, na.rm = TRUE),
    Min = min(EstimatedSalary, na.rm = TRUE),
    Max = max(EstimatedSalary, na.rm = TRUE)
  )

summary_female_salary <- bank_dataset %>%
  filter(Gender == "Female") %>%
  summarise(
    Median = median(EstimatedSalary, na.rm = TRUE),
    Q1 = quantile(EstimatedSalary, 0.25, na.rm = TRUE),
    Q3 = quantile(EstimatedSalary, 0.75, na.rm = TRUE),
    Min = min(EstimatedSalary, na.rm = TRUE),
    Max = max(EstimatedSalary, na.rm = TRUE)
  )

# Print summary statistics
cat("Summary Statistics for Female Salary:\n")
print(summary_female_salary)
cat("\nSummary Statistics for Male Salary:\n")
print(summary_male_salary)
```

Conclusion: The box plot illustrates the distribution of estimated salaries between different genders. There appear to be minimal differences between genders in terms of salary distribution.

